Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-23 paper/zhang-cook-laa-langrene-menendez.Rmd
Small changes in application section.

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. For SA, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for CRS. The original trace plot for a 2D projection problem, optimised by CRS, is shown on the left panel of Figure \ref{fig:interruption} and one can observe clearly that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at Time 61 has a higher index value than the target in the third interpolation. Rather than starting the next iteration from the target basis on Time 65, CRS starts the next iteration at Time 61 on the right panel and reaches a better final basis.

@@

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to see if an even better basis can be found. A polish search takes the final basis of an optimiser as the start of a new guided tour to search for local breakthrough. The polish algorithm is similar to the CRS but with three distinctions: 1) a hundred rather than one candidate bases are generated each time in the inner loop; 2) the neighbourhood size is reduced in the inner loop, rather than in the outer loop; and 3) three more termination conditions have been added to ensure the new basis generated is distinguishable from the current one in terms of the distance in the space,  percentage change in the index value, and neighbourhood size:

@@

Figure \ref{fig:polish} presents the projectd data and trace plot of a 2D projection, optimised by CRS and followed by the polish. In the top row shows the intial projection, the final projection after CRS, and the final projection after polish, repsectively. The end basis found by CRS reveals the four clusters in the data, but the edges of each cluster are not clean-cut. Polish works with this end basis and further pushes the index value to produce much clear edges of the cluster, especially along the vertical axis.

@@

The holes index function used for all the examples before this section produces a smooth interpolation, while this is not the case for all the indices. A 1D projection function, $I^{K}(n)$, compar‚es the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, based on the Kolmogorov test. Let $F_{.}(n)$ be the ECDF function, with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov index, $I^{nk}(n)$, is defined as:

@@

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While CRS and SA have managed to get close to the index value of the theoretical best, trace plot shows that it takes SA much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SA tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. Pseudo Derivative (PD) quickly gets stuck near the starting position. Comparing the amount of random search done by CRS and SA, it is surprising that SA does not carry as many samples as CRS. Combining the insights from both the trace and PCA plot, one can learn the two different search strategies by CRS and SA: CRS frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SA first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as what SA has done, can lead to a more efficient search, in terms of the overall number of points evaluated.

@@

The next experiment compares the performance of CRS and SA when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With CRS and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing CRS and SA for a search size of 0.5, SA does not seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SA is working hard to examine if there is any other optimum in the basis space but the relative small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SA, with a reasonable large search size, is able to overcome the local optimum and optimise close towards the global optimum.

>>>>

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. For SA, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for CRS. The original trace plot for a 2D projection problem, optimised by CRS, is shown on the left panel of Figure \ref{fig:interruption} and one can observe that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at Time 61 has a higher index value than the target in the third interpolation. Rather than starting the next iteration from the target basis on Time 65, CRS starts the next iteration at Time 61 on the right panel and reaches a better final basis.

@@

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to see if an even better basis can be found. A polish search takes the final basis of an optimiser as the start of a new guided tour to search for local improvements. The polish algorithm is similar to the CRS but with three distinctions: 1) a hundred rather than one candidate bases are generated each time in the inner loop; 2) the neighbourhood size is reduced in the inner loop, rather than in the outer loop; and 3) three more termination conditions have been added to ensure the new basis generated is distinguishable from the current one in terms of the distance in the space,  percentage change in the index value, and neighbourhood size:

@@

Figure \ref{fig:polish} presents the projectd data and trace plot of a 2D projection, optimised by CRS and followed by the polish step. The top row shows the intial projection, the final projection after CRS, and the final projection after polish, repsectively. The end basis found by CRS reveals the four clusters in the data, but the edges of each cluster are not clean-cut. Polish works with this end basis and further pushes the index value to produce much clear edges of the cluster, especially along the vertical axis.

@@

The holes index function used for all the examples before this section produces a smooth interpolation, while this is not the case for all the indices. An example of a noisy index function for 1D projections compares the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, using the Kolmogorov test. Let $F_{.}(n)$ be the ECDF function, with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov index, $I^{nk}(n)$, is defined as:

@@

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated paths are smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index functions. While CRS and SA have managed to get close to the index value of the theoretical best, the trace plot shows that it takes SA much longer to find the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SA tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. Pseudo Derivative (PD) quickly gets stuck near the starting position. Comparing the amount of random search done by CRS and SA, it is surprising that SA does not carry as many samples as CRS. Combining the insights from both the trace and PCA plot, one can learn the two different search strategies by CRS and SA: CRS frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SA first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as SA has done here, can lead to a more efficient search, in terms of the overall number of points evaluated.

@@

The next experiment compares the performance of CRS and SA when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With CRS and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing CRS and SA for a search size of 0.5, SA does not seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SA is working hard to examine if there is any other optimum in the basis space but the relative small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SA, with a reasonable large search window, is able to overcome the local optimum and optimise close towards the global optimum.

### 2021-03-23 paper/zhang-cook-laa-langrene-menendez.tex
Small changes in application section.

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projection
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and this workflow can be applied to
a broad class of optimisers, to assess their performance. An R package,
\CRANpkg{ferrn} \citep{ferrn}, has been created to implement the
diagnostics.

@@

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data
structure is defined to capture information during the optimisation, and
different diagnostic plots are designed with the data collected. Section
\ref{application} shows applications of how these diagnostic plots can
be used to discover interesting aspects of different optimisers and
guide improvements to the existing algorithms. Finally, Section
\ref{implementation} describes the R package: \CRANpkg{ferrn}
\citep{ferrn}, that implements the visual diagnostics.

@@

{\centering \includegraphics[width=0.5\linewidth,height=0.2\textheight]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/img/tour-path}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-torus-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1}

@@

The holes index function used for all the examples before this section
produces a smooth interpolation, while this is not the case for all the
indices. A 1D projection function, \(I^{K}(n)\), compar‚es the projected
data, \(\mathbf{Y}_{n \times 1}\), to a randomly generated normal
distribution, \(\mathcal{N}_{n \times 1}\), based on the Kolmogorov
test. Let \(F_{.}(n)\) be the ECDF function, with two possible
subscripts \(Y\) and \(\mathcal{N}\) representing the projected and
randomly generated data, and \(n\) denoting the number of observation,
the Normal Kolmogorov index, \(I^{nk}(n)\), is defined as:

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-result-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/flip-sign-1}

@@

This project contributes to the software development in two packages: a
data collection object is implemented in \CRANpkg{tourr} \citep{tourr},
while the visual diagnostics of the optimisers is implemented in
\CRANpkg{ferrn} \citep{ferrn}. Below list functions in the
\CRANpkg{ferrn} \citep{ferrn} package:

>>>>

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projection
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and this workflow can be applied to
a broad class of optimisers, to assess their performance. An R package,
\pkg{ferrn}, has been created to implement the diagnostics.

@@

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data
structure is defined to capture information during the optimisation, and
different diagnostic plots are designed with the data collected. Section
\ref{application} shows applications of how these diagnostic plots can
be used to discover interesting aspects of different optimisers and
guide improvements to the existing algorithms. Finally, Section
\ref{implementation} describes the R package: \pkg{ferrn}, that
implements the visual diagnostics.

@@

{\centering \includegraphics[width=0.5\linewidth,height=0.2\textheight]{/Users/ursulalaa/papers/paper-tour-vis/img/tour-path}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/toy-search-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/toy-interp-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/toy-pca-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/toy-pca-animated-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/toy-tour-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/toy-torus-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/polish-1}

@@

The holes index function used for all the examples before this section
produces a smooth interpolation, while this is not the case for all the
indices. An example of a noisy index function for 1D projections
compares the projected data, \(\mathbf{Y}_{n \times 1}\), to a randomly
generated normal distribution, \(\mathcal{N}_{n \times 1}\), using the
Kolmogorov test. Let \(F_{.}(n)\) be the ECDF function, with two
possible subscripts \(Y\) and \(\mathcal{N}\) representing the projected
and randomly generated data, and \(n\) denoting the number of
observation, the Normal Kolmogorov index, \(I^{nk}(n)\), is defined as:

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/noisy-better-geo-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/kol-result-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-tour-vis/figs/flip-sign-1}

@@

The implementation of this projection has been divided into two
packages: the data collection object is implemented in the existing CRAN
package \CRANpkg{tourr} \citep{tourr} while the optimiser diagnostics
have been implemented in a new package, \pkg{ferrn}. When a guided tour
is run, the users can choose to collect the data by binding
\texttt{animate\_*()} to a variable. Once the data object has been
obtained, the \pkg{ferrn} package provides diagnostic plots shown in
Section \ref{vis-diag}. The main functions of the \pkg{ferrn} package
functionality is listed below.
